### 2016-07-18 plosOneJordanousKeller/PAPER/plos_JordanousKeller.tex

NEED TO RE-WRITE THIS EVENTUALLY:
{\em
Creativity is a complex multi-dimensional concept which resists definition; it is accessible to us at a subjective, multi-faceted and dynamic level rather than being fixed and clearly defined. If we wish to study creativity scientifically using computational models, or simulate it computationally, then a tractable and accurate interpretation of creativity is needed. Creativity encompasses many related aspects, abilities, properties and behaviours. This paper uses techniques from the field of statistical natural language processing to identify a collection of fourteen key components of creativity through an analysis of what is considered to be important in talking about creativity. Words were identified which appeared significantly often in connection with discussions of the concept. Using a measure of lexical similarity to help cluster these words, a number of distinct themes emerged, which collectively contribute to a comprehensive and multi-perspective definition of creativity. The components provide an ontology of creativity: a set of building blocks which, in evaluation, have proven useful for making the concept easier to understand and more tractable to study and evaluate.}

@@

The aim of the work reported in this paper is to examine the nature of creativity and to identify within it a set of components, representing key dimensions, that are recognised across a combination of different viewpoints. We present a novel, empirical approach to this problem that focuses on what is revealed about our understanding of creativity and its attributes by the words we use to discuss and debate the nature of the concept. Analysis of this language provides a sound basis for constructing a sufficiently detailed and comprehensive model of creativity \cite{lakoff87,wittgenstein58}. The current work is intended as a significant, methodological contribution towards addressing the Grand Challenge of evaluation in computational creativity research. It should provide researchers with a firm foundation for evaluating exactly how creative so-called {\emph creative systems} actually are.

In our approach, statistical language processing techniques are used to identify words significantly associated with creativity in a corpus of academic papers on the subject. A corpus spanning some 60 years of research into the nature of creativity was collected together. The papers were gathered from a wide variety of disciplines including psychology, educational testing and computational creativity, amongst others. The language data drawn from this collection was then analysed and contrasted with data from a corpus of matched papers on subjects unrelated to creativity. From this analysis, a set of 694 {\em creativity words\/} was identified, where each creativity word appeared significantly more often than expected in the corpus of creativity papers. A measure of lexical similarity provided a basis for clustering the creativity words into groups of words with similar or shared aspects of meaning. Through inspection of these clusters, a total of fourteen {\em key components of creativity\/} was identified, where each represents a key theme or aspect of creativity. The set of components yields information about the nature of creativity, based on what is collectively emphasised in discussions about the concept.

@@

In the rest of this section we begin by noting a variety of attempts to define creativity. The representation of subjective, ambiguous, loosely structured concepts is also considered.
%, both from the perspective of current Semantic Web research and from a broader conceptual perspective. 
In the remaining sections, details are provided of the methodology used to identify components of creativity from an analysis of language data. The results of this analysis are then presented in terms of a model that encompasses fourteen key components or dimensions of creativity. The derived set of components is evaluated in terms of how well it satisfies the need for a shared, inclusive and comprehensive account of creativity and provides a vocabulary of creativity that is accessible to both people and machines. Finally, conclusions are drawn and some directions for further work are outlined.

@@

\noindent In short, we need to specify and justify the standards that we use to judge creativity. A more objective and well-articulated definition of creativity enables researchers to make a worthwhile contribution \cite{torrance67,plucker04defn,kaufman09}.

@@

\subsection*{Towards a model of creativity}\label{progressingForwards}

@@

Each word was assigned a part-of-speech tag identifying its grammatical category (i.e. whether the word was a noun, verb, preposition, etc.). Such tagging is useful because it allows us to distinguish between different grammatical uses of a common orthographic form. For example, the use of {\em novel\/} as a noun in {\em a good novel\/} can be properly differentiated from its use as an adjective in {\em a novel idea\/}.  The data was further simplified and filtered so that only words of the four `major' categories (i.e. noun, verb, adjective and adverb) were represented. Note that the major categories bear the semantic content of the papers making up the creativity corpus. They may be distinguished from minor categories or `function words', such as pronouns ({\em something\/}, {\em itself\/}) prepositions (e.g. {\em upon\/}, {\em by\/}) conjunctions ({\em but\/}, {\em or\/}) and quantifiers (e.g. {\em many\/}, {\em more\/}). Such words have little independent semantic content so are of limited interest for the present study and may be removed from the data.

@@

The word frequency data derived from the two corpora was used to establish which words occur significantly more often in the creativity corpus than in the non-creativity corpus. This in turn can be regarded as providing evidence of which words are salient to the definition of creativity. Salient words were identified using the log-likelihood ratio  (also referred to as the $G^2$ or G-squared statistic), which is a measure of how well observed data fits a model or expected distribution \cite{dunning93,kilgarriff01,rayson00,oakes98}. It provides an alternative to Pearson's chi-squared ($\chi^2$) test and has been advocated as the more appropriate measure of the two for corpus analysis as it does not rely on the (unjustifiable) assumption of normality in word distribution  \cite{dunning93,kilgarriff01,oakes98}. This is a particular issue when analysing smaller corpora, such as those used in the present work (see Fig. 1). %\ref{S1_Fig}). 
The log likelihood ratio statistic is more accurate in its treatment of infrequent words in the data,  which often hold useful information. By contrast, the $\chi^2$ statistic tends to under-emphasise such outliers at the expense of very frequently occurring data points.

@@

The creativity words were clustered according a statistical measure of {\em distributional similarity\/}. The intuition underlying distributional measures of similarity derives from the {\em distributional hypothesis\/} due to Harris \cite{harris68}. This hypothesis  states that similarity of distribution correlates with similarity of meaning: two words that tend to appear in similar linguistic contexts will tend to have similar meanings. The notion of linguistic context here is not fixed and might plausibly be modelled in a variety of different ways. For example, two words might be considered to inhabit the same context if they appear in the same document or the same sentence or if they stand in the same grammatical relationship to some other word (e.g. both occur as {\em subject\/} of a particular verb or {\em modifier\/} of a given noun). In practice it has been shown that modelling distribution in terms of grammatical relations leads to a tighter correlation between distributional similarity and closeness of meaning \cite{kilgarriff00}.

@@

\noindent
Grammatical relations were obtained from an analysis of the written portion of the British National Corpus \cite{leech92}, which had previously been processed using the RASP toolkit \cite{briscoe06} in order to extract them. Using this data, each word in the creativity corpus was associated with a list of all of the grammatical relations with which it occurred, together with their corresponding counts of occurrence.\footnote{Not all of the grammatical relation information output by RASP was used to calculate distributional similarity. In practice, just the subject, object and modifier relation types are used as these tend to give the best results \cite{weeds03}.} A potential difficulty with obtaining word similarity data based on the BNC (i.e. using data from sources of everyday usage of English, rather than from more specialist sources) would arise if the majority of the creativity words were used with distinctive or technical senses within the creativity corpus. From inspection and from knowledge of creativity literature, however, this situation was found to be unlikely. While some narrowly specialised usage may be present to some small degree in the set of creativity words, most words retain general senses as reflected in the wider BNC data set. An advantage of unsing the BNC is that its  size increases the chances of a comprehensive coverage of the general senses of each word of interest.

@@

As part of the manual inspection process, candidate components were further considered in terms of the {\em Four Ps} of creativity \cite{rhodes61,mooney63,mackinnon70,kaufman09} described earlier in this paper. This additional analysis provided a means of identifying alternative perspectives and revealing subtle (but still important) aspects of creativity.  For example, {\em novelty} is commonly associated with the results of creative behaviour ({\em product\/}): how novel is the artefact or idea that has been produced? However, we could similarly recognise as creative an approach to a task  ({\em process\/}) that does things in a novel and different way.  Also, if a product is new in a particular environment ({\em press\/}), then it may well be be regarded as creative to those in that environment. Viewing {\em novelty} from the perspectives of {\em product}, {\em process} and {\em press} uncovers these subtle and interlinked distinctions.

@@

From the analysis steps described in the previous section it was possible to extract a set of fourteen key components of creativity. These components are summarised in Fig. 5 %\ref{S5_Fig} 
and are presented in more detail below. The components contribute collectively to the overall concept and may be regarded as providing an {\em ontology of creativity\/}.  It is important to note, however that the fourteen components do not constitute a necessary and sufficient definition of creativity, in all its possible manifestations. There are two reasons for this. Firstly, some of the components we have identified appear to be logically inconsistent with others in the set. Consider for example the apparent need for autonomous, independent behaviour identified in {\em Independence and Freedom\/} and contrast this with the requirement for social interaction implied by {\em Social Interaction and Communication\/}. %\hidden{These factors may appear logically opposed.}
Secondly, of course, creativity %\hidden{does not lend itself to formal and precise definition and 
%I added this bit to clarify the point a bit [does not lend itself to formal and precise definition and]
also manifests itself in rather different ways across different domains \cite{plucker04} and components will vary in importance, according to the requirements of a particular domain. As an illustration of this second point, %Can we explicitly relate this illustration to one or more of the components that have been identified?
creative behaviour in mathematical reasoning has more focus on finding a correct solution to a problem than is the case for creative behaviour in, say,  musical improvisation \cite{colton08,jordanous12jims}.

@@

The fourteen components provide a fuller and clearer account of the constituent parts of the concept of creativity. An important aim of the current work is to make the components available as a resource for other researchers in computational creativity and to provide a basis for the automated evaluation of creative systems. As a step in this direction, the components have been expressed in an open, machine-readable form within the Semantic Web. In this way, the definition of the components benefits from and are enriched by concepts that are already defined within the Semantic Web. %To achieve this, SKOS (Simple Knowledge Organisation System) \cite{skos} was used as a basis for constructing an OWL ontology representation of creativity and its components,

@@

In both case studies, the components of creativity were chosen as the way of defining creativity for step 1a of SPECS, and were weighted according to their importance and relevance for creativity in the creative domains under study for each case study (step 1b of SPECS). Each component was treated as one standard to be used to evaluate the creativity of the creative systems in the case studies (step 2 of SPECS). Each case study system was then tested against each component using feedback provided by judges (step 3 of SPECS), resulting in a detailed set of evaluative feedback on the creativity of each system in the case studies.

@@

The resulting component-based evaluation yielded detailed information about creative strengths and weaknesses of the systems under investigation, highlighting those components where a system performs strongly. Crucially, the evaluation feedback also highlighted areas where a given system performed poorly. For example, in the musical improvisation study, Case Study 1, we found that all systems' creativity could be improved most by improving performance in {\em Social Interaction and Communication}, {\em Intention and Emotional Involvement} and {\em Domain Competence} (the three components found to be most important for creativity in musical improvisation). Similarly, it is useful to be able to quickly obtain formative feedback on strengths and weaknesses in time-limited scenarios such as that replicated in Case Study 2 during the development of creative systems (when ongoing evaluation of progress ideally needs to be both timely and time-efficient). Insight can then be obtained on where future development effort is best spent.

>>>>

\sout{
Creativity is a complex multi-dimensional concept which resists definition; it is accessible to us at a subjective, multi-faceted and dynamic level rather than being fixed and clearly defined. If we wish to study creativity scientifically using computational models, or simulate it computationally, then a tractable and accurate interpretation of creativity is needed. Creativity encompasses many related aspects, abilities, properties and behaviours. This paper uses techniques from the field of statistical natural language processing to identify a collection of fourteen key components of creativity through an analysis of what is considered to be important in talking about creativity. Words were identified which appeared significantly often in connection with discussions of the concept. Using a measure of lexical similarity to help cluster these words, a number of distinct themes emerged, which collectively contribute to a comprehensive and multi-perspective definition of creativity. The components provide an ontology of creativity: a set of building blocks which, in evaluation, have proven useful for making the concept easier to understand and more tractable to study and evaluate.}

Creativity is a complex, multi-faceted concept encompassing a variety of related aspects, abilities, properties and behaviours. If we wish to study creativity scientifically, then a tractable and well-articulated model of how creativity behaviour emerges is required. Such a model would be of great value to researchers investigating the nature of creativity and in particular, those concerned with the evaluation of creative practice. This paper describes a unique approach to developing a suitable model of how creative behaviour emerges that is based on the words people used to describe the concept. Using techniques from the field of statistical natural language processing, we identify a collection of fourteen key components of creativity through an analysis of a corpus of academic papers on the topic. Words are identified which appear significantly often in connection with discussions of the concept. Using a measure of lexical similarity to help cluster these words, a number of distinct themes emerge, which collectively contribute to a comprehensive and multi-perspective model of creativity. The components provide an ontology of creativity: a set of building blocks which can be used to model creative practice in a variety of domains. The components have been employed in two case studies to evaluate the creativity of computational systems and have proven useful in articulating the strengths and weaknesses of computational creativity systems.

@@

The aim of the work reported in this paper is to examine the nature of creativity and to identify within it a set of components, representing key dimensions, that are recognised across a combination of different viewpoints. We present a novel, empirical approach to the problem of modelling how creative behaviour emerges, that focuses on what is revealed about our understanding of creativity and its attributes by the words we use to discuss and debate the nature of the concept. Analysis of this language provides a sound basis for constructing a sufficiently detailed and comprehensive model of creativity \cite{lakoff87,wittgenstein58}. The current work is intended as a significant, methodological contribution towards addressing the Grand Challenge of evaluation in computational creativity research. It should provide researchers with a firm foundation for evaluating exactly how creative so-called {\emph creative systems} actually are.

On our approach, statistical language processing techniques are used to identify words significantly associated with creativity in a corpus of academic papers on the subject. A corpus spanning some 60 years of research into the nature of creativity was collected together. The papers were gathered from a wide variety of disciplines including psychology, educational testing and computational creativity, amongst others. The language data drawn from this collection was then analysed and contrasted with data from a corpus of matched papers on subjects unrelated to creativity. From this analysis, a set of 694 {\em creativity words\/} was identified, where each creativity word appeared significantly more often than expected in the corpus of creativity papers. A measure of lexical similarity provided a basis for clustering the creativity words into groups of words with similar or shared aspects of meaning. Through inspection of these clusters, a total of fourteen {\em key components of creativity\/} was identified, where each represents a key theme or attribute of creativity. The set of components yields information about the nature of creativity, based on what is collectively emphasised in discussions about the concept.

@@

In the rest of this section we begin by noting a variety of attempts to define creativity. The representation of subjective, ambiguous, loosely structured concepts is considered.
%, both from the perspective of current Semantic Web research and from a broader conceptual perspective. 
In the remaining sections, details are provided of the methodology used to identify components of creativity from an analysis of language data. The results of this analysis are then presented in terms of a model that encompasses fourteen key components. The derived set of components is evaluated in terms of how well it satisfies the need for a shared, inclusive and comprehensive account of creativity and provides a vocabulary of creativity that is accessible to both people and machines. Finally, conclusions are drawn and some directions for further work are outlined.

@@

\noindent In short, we need to specify and justify the standards that we use to judge creativity. A more objective and well-articulated account of how creativity is manifested enables researchers to make a worthwhile contribution \cite{torrance67,plucker04defn,kaufman09}.

@@

%\subsection*{Towards a model of creativity}\label{progressingForwards}

@@

Each word was assigned a part-of-speech tag identifying its grammatical category (i.e. whether the word was a noun, verb, preposition, etc.). Such tagging is useful because it allows us to distinguish between different grammatical uses of a common orthographic form. For example, the use of {\em novel\/} as a noun in {\em a good novel\/} can be properly differentiated from its use as an adjective in {\em a novel idea\/}.  The data was further simplified and filtered so that only words of the four `major' categories (i.e. noun, verb, adjective and adverb) were represented. Note that the major categories bear the semantic content of the papers making up the creativity corpus. They may be distinguished from minor categories or `function words', such as pronouns ({\em something\/}, {\em itself\/}) prepositions (e.g. {\em upon\/}, {\em by\/}) conjunctions ({\em but\/}, {\em or\/}) and quantifiers (e.g. {\em many\/}, {\em more\/}). Because such words have little independent semantic content, they are of limited interest for the present study and may be removed from the data.

@@

The word frequency data derived from the two corpora was used to establish which words occur significantly more often in the creativity corpus than in the non-creativity corpus. This in turn can be regarded as providing evidence of which words are salient to the definition of creativity. Salient words were identified using the log-likelihood ratio  (also referred to as the $G^2$ or G-squared statistic), which is a measure of how well observed data fit a model or expected distribution \cite{dunning93,kilgarriff01,rayson00,oakes98}. It provides an alternative to Pearson's chi-squared ($\chi^2$) test and has been advocated as the more appropriate measure of the two for corpus analysis as it does not rely on the (unjustifiable) assumption of normality in word distribution  \cite{dunning93,kilgarriff01,oakes98}. This is a particular issue when analysing smaller corpora, such as those used in the present work (see Fig. 1). %\ref{S1_Fig}). 
The log likelihood ratio statistic is more accurate in its treatment of infrequent words in the data,  which often hold useful information. By contrast, the $\chi^2$ statistic tends to under-emphasise such outliers at the expense of very frequently occurring data points.

@@

\sout{The creativity words were clustered according a statistical measure of {\em distributional similarity\/}. }The intuition underlying distributional measures of similarity derives from the {\em distributional hypothesis\/} due to Harris \cite{harris68}. This hypothesis  states that similarity of distribution correlates with similarity of meaning: two words that tend to appear in similar linguistic contexts will tend to have similar meanings. The notion of linguistic context here is not fixed and might plausibly be modelled in a variety of different ways. For example, two words might be considered to inhabit the same context if they appear in the same document or the same sentence or if they stand in the same grammatical relationship to some other word (e.g. both occur as {\em subject\/} of a particular verb or {\em modifier\/} of a given noun). In practice it has been shown that modelling distribution in terms of grammatical relations leads to a tighter correlation between distributional similarity and closeness of meaning \cite{kilgarriff00}.

@@

\noindent
Grammatical relations were obtained from an analysis of the written portion of the British National Corpus \cite{leech92}, which had previously been processed using the RASP toolkit \cite{briscoe06} in order to extract them. Using this data, each word in the creativity corpus was associated with a list of all of the grammatical relations with which it occurred, together with their corresponding counts of occurrence.\footnote{Not all of the grammatical relation information output by RASP was used to calculate distributional similarity. In practice, just the subject, object and modifier relation types are used as these tend to give the best results \cite{weeds03}.} A potential difficulty with obtaining word similarity data based on the BNC (i.e. using data from sources of everyday usage of English, rather than from more specialist sources) would arise if the majority of the creativity words were used with distinctive or technical senses within the creativity corpus. From inspection and from knowledge of creativity literature, however, this situation was found to be unlikely. While some narrowly specialised usage may be present to some small degree in the set of creativity words, most words retain general senses as reflected in the wider BNC data set. An advantage of using the BNC is that its  size increases the chances of a comprehensive coverage of the general senses of each word of interest.

@@

As part of the manual inspection process, candidate components were further considered in terms of the {\em Four Ps} of creativity \cite{rhodes61,mooney63,mackinnon70,kaufman09} described earlier in this paper. This additional analysis provided a means of identifying alternative perspectives and revealing subtle (but still important) aspects of creativity.  For example, {\em novelty} is commonly associated with the results of creative behaviour ({\em product\/}): how novel is the artefact or idea that has been produced? However, we could similarly recognise as creative an approach to a task  ({\em process\/}) that does things in a novel and different way.  Also, if a product is new in a particular environment ({\em press\/}), then it may well be regarded as creative to those in that environment. Viewing {\em novelty} from the perspectives of {\em product}, {\em process} and {\em press} uncovers these subtle and interlinked distinctions.

@@

From the analysis steps described in the previous section it was possible to extract a set of fourteen key components of creativity. These components are summarised in Fig. 5 %\ref{S5_Fig} 
and are presented in more detail below. The components contribute collectively to the overall concept and may be regarded as providing an {\em ontology of creativity\/}.  It is important to note, however that the fourteen components do not constitute a set of necessary and sufficient conditions for creativity, in all its possible manifestations. There are two reasons for this. Firstly, some of the components we have identified appear to be logically inconsistent with others in the set. Consider for example the apparent need for autonomous, independent behaviour identified in {\em Independence and Freedom\/} and contrast this with the requirement for social interaction implied by {\em Social Interaction and Communication\/}. %\hidden{These factors may appear logically opposed.}
Secondly, of course, creativity %\hidden{does not lend itself to formal and precise definition and 
%I added this bit to clarify the point a bit [does not lend itself to formal and precise definition and]
also manifests itself in rather different ways across different domains \cite{plucker04} and components will vary in importance, according to the requirements of a particular domain. As an illustration of this second point, %Can we explicitly relate this illustration to one or more of the components that have been identified?
creative behaviour in mathematical reasoning has more focus on finding a correct solution to a problem than is the case for creative behaviour in, say,  musical improvisation \cite{colton08,jordanous12jims}.

@@

The fourteen components provide a fuller and clearer account of the constituent parts of the concept of creativity. An important aim of the current work is to make the components available as a resource for other researchers in computational creativity and to provide a basis for the automated evaluation of creative systems. As a step in this direction, the components have been expressed in an open, machine-readable form within the Semantic Web. In this way, the characterization of the components benefits from and is enriched by concepts that are already represented within the Semantic Web. %To achieve this, SKOS (Simple Knowledge Organisation System) \cite{skos} was used as a basis for constructing an OWL ontology representation of creativity and its components,

@@

In both case studies, the components of creativity were chosen as the way of characterising creativity for step 1a of SPECS, and were weighted according to their importance and relevance for creativity in the creative domains under study for each case study (step 1b of SPECS). Each component was treated as one standard to be used to evaluate the creativity of the creative systems in the case studies (step 2 of SPECS). Each case study system was then tested against each component using feedback provided by judges (step 3 of SPECS), resulting in a detailed set of evaluative feedback on the creativity of each system in the case studies.

@@

The resulting component-based evaluation yielded detailed information about creative strengths and weaknesses of the systems under investigation, highlighting those components where a system performs strongly. Crucially, the evaluation feedback also highlighted areas where a given system performed poorly. For example, in the musical improvisation study, Case Study 1, we found that, in general, creativity could be improved most by improving performance in {\em Social Interaction and Communication}, {\em Intention and Emotional Involvement} and {\em Domain Competence} (the three components found to be most important for creativity in musical improvisation). Similarly, it is useful to be able to quickly obtain formative feedback on strengths and weaknesses in time-limited scenarios such as that replicated in Case Study 2 during the development of creative systems (when ongoing evaluation of progress ideally needs to be both timely and time-efficient). Insight can then be obtained on where future development effort is best spent.
